Final project

### 2025-04-05 products/manuscript/Manuscript.qmd
Changes

# Load necessary libraries

```{r}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(vcd)
library(caret)        # For data partitioning and model training
library(randomForest) # For Random Forest modeling
library(tidyverse)
library(tidymodels)
library(broom)
library(here)           # For file paths
library(pROC)
library(naniar)         # For gg_miss_var()
```

# PART 1: DATA IMPORT AND CLEANING

```{r}
# Import the dataset
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

@@

# Create histograms and density overlays for key variables

# Age distribution
ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()

# Serum Cholesterol distribution (excluding zeros)
ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()

# Resting Blood Pressure distribution (excluding zeros)
ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

# Boxplot for resting blood pressure by heart disease status
ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
  geom_boxplot() +
  labs(title = "Resting Blood Pressure by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Resting Blood Pressure (mmHg)") +
  theme_minimal()

# Violin plot for oldpeak by heart disease status
ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(trim = FALSE) +
  labs(title = "Oldpeak Distribution by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Oldpeak") +
  theme_minimal()

@@

# Plot the correlation matrix with ggcorrplot
ggcorrplot(cor_matrix, 
           lab = TRUE,                         
           lab_size = 3,                       
           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
           title = "Correlation Matrix", 
           ggtheme = theme_minimal())       
```

@@

@@

@@

heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = count), color = "black", size = 5) +
  facet_wrap(~ target, ncol = 2) +
  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
  labs(title = "Interaction between Blood Pressure and ST Depression",
       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "ST Depression Status",
       fill = "Count") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
print(heatmap_plot)

# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

>>>>

# Load necessary libraries
```{r}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(vcd)
library(caret)        # For data partitioning and model training
library(randomForest) # For Random Forest modeling
library(tidyverse)
library(tidymodels)
library(broom)
library(here)           # For file paths
library(pROC)
library(naniar)         # For gg_miss_var()
```

# PART 1: DATA IMPORT AND CLEANING
```{r}
# Import the dataset
cardio_df <- read_csv(here("data", "raw-data", "Cardiovascular_Disease_Dataset.csv"))

@@

summary_df = skimr::skim(summary_stats)
print(summary_stats)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)

# Age distribution
p1 <- ggplot(cardio_df, aes(x = age)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "steelblue", alpha = 0.5) +
  geom_density(color = "darkblue", linewidth = 1) +
  labs(title = "Age Distribution", x = "Age", y = "Density") +
  scale_x_continuous(breaks = seq(0, max(cardio_df$age, na.rm = TRUE) + 10, by = 10)) +
  theme_bw()

plot(p1)
figure_file = here("results", "figures","Age distribution.png")
ggsave(filename = figure_file, plot=p1)

# Serum Cholesterol distribution (excluding zeros)
p2 <- ggplot(subset(cardio_df, serumcholestrol > 0), aes(x = serumcholestrol)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "forestgreen", alpha = 0.5) +
  geom_density(color = "darkgreen", linewidth = 1) +
  labs(title = "Serum Cholesterol Distribution", 
       x = "Serum Cholesterol (mg/dL)", y = "Density") +
  theme_minimal()

plot(p2)
figure_file = here("results", "figures","Serum Cholestrol.png")
ggsave(filename = figure_file, plot=p2)

# Resting Blood Pressure distribution (excluding zeros)
p3 <- ggplot(subset(cardio_df, restingBP > 0), aes(x = restingBP)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "maroon", alpha = 0.5) +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Resting Blood Pressure Distribution", 
       x = "Resting Blood Pressure (mmHG)", y = "Density") +
  theme_bw()

plot(p3)
figure_file = here("results", "figures","Resting Blood Pressure Distribution.png")
ggsave(filename = figure_file, plot=p3)

# Boxplot for resting blood pressure by heart disease status
p4 <- ggplot(cardio_df, aes(x = target, y = restingBP, fill = target)) +
  geom_boxplot() +
  labs(title = "Resting Blood Pressure by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Resting Blood Pressure (mmHg)") +
  theme_minimal()

plot(p4)
figure_file = here("results", "figures","restingBP by heart disease status.png")
ggsave(filename = figure_file, plot=p4)

# Violin plot for oldpeak by heart disease status
p5 <- ggplot(cardio_df, aes(x = target, y = oldpeak, fill = target)) +
  geom_violin(trim = FALSE) +
  labs(title = "Oldpeak Distribution by Heart Disease Status",
       x = "Heart Disease Status",
       y = "Oldpeak") +
  theme_minimal()

plot(p5)
figure_file = here("results", "figures","oldpeak by heart disease status.png")
ggsave(filename = figure_file, plot=p5)

@@

# Plot the correlation matrix with ggcorrplot
p6 <- ggcorrplot(cor_matrix, 
           lab = TRUE,                         
           lab_size = 3,                       
           colors = c("blue", "white", "red"),  # Color gradient from blue (negative) to red (positive)
           title = "Correlation Matrix", 
           ggtheme = theme_minimal())

plot(p6)
figure_file = here("results", "figures","correlation matrix.png")
ggsave(filename = figure_file, plot=p6)     
```

@@

# Save the processed data
saveRDS(cleaned_df, file = "../../data/processed-data/cleaned_df.rds")

@@

summary_df = skimr::skim(summary_table)
print(summary_table)
# save to file
summarytable_file = here("results", "tables", "summarytable2.rds")
saveRDS(summary_df, file = summarytable_file)

@@

p7 <- heatmap_plot <- ggplot(interaction_counts, aes(x = elevated_BP, y = elevated_oldpeak, fill = count)) +
  geom_tile(color = "white", size = 0.5) +
  geom_text(aes(label = count), color = "black", size = 5) +
  facet_wrap(~ target, ncol = 2) +
  scale_fill_gradient(low = "lightgreen", high = "forestgreen") +
  labs(title = "Interaction between Blood Pressure and ST Depression",
       subtitle = "By Heart Disease Status (Individuals <50 with Normal Cholesterol)",
       x = "Blood Pressure Status",
       y = "ST Depression Status",
       fill = "Count") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

plot(p7)
figure_file = here("results", "figures","interaction heatmap.png")
ggsave(filename = figure_file, plot=p7)

# Visualize CVD risk by blood pressure and ST depression status using faceted grouped Bar Chart
p8 <- ggplot(interaction_counts, aes(x = elevated_BP, y = count, fill = target)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  facet_wrap(~ elevated_oldpeak, ncol = 2, scales = "free_y") +
  labs(title = "CVD Risk by Blood Pressure and ST Depression Status",
       subtitle = "Individuals Under 50 with Normal Cholesterol (<200 mg/dL)",
       x = "Blood Pressure Status",
       y = "Count",
       fill = "Heart Disease") +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

plot(p8)
figure_file = here("results", "figures","CVD risk by BP and ST depression.png")
ggsave(filename = figure_file, plot=p8)
```
